---
title: "08E - Programming: Introduction to Functions in R"
author: "Piotr Kosowski"
date: "2025-03-08"
---

# Introduction to Functions

**Functions** are one of the most essential building blocks in R programming. They allow you to **encapsulate** reusable blocks of code, increasing:

-   **Efficiency**: Write once, reuse many times\
-   **Modularity**: Break complex tasks into smaller pieces\
-   **Clarity**: Keep your scripts organized and easier to maintain

A function in R can accept **arguments** (inputs) and can optionally **return** a value (output). This flexibility enables you to automate repetitive tasks and design cleaner, more maintainable code.

------------------------------------------------------------------------

# 1. What is a Function in R?

A function is a named object that **performs a specific task**. R has many built-in functions (e.g., `mean()`, `sum()`, etc.), but you can also create your own.

**General syntax** for defining a function:

``` r
name_of_function <- function(arg1, arg2, ...) {
  # Code to execute when the function is called
  # optionally: return(some_value)
}
```

-   **`name_of_function`**: a descriptive identifier\
-   **`arg1`, `arg2`, ...**: parameters/arguments\
-   **Code block**: The statements that run whenever you call the function, enclosed in `{ ... }`.

------------------------------------------------------------------------

# 2. Creating Functions in R

We’ll walk through multiple examples to show how functions are defined and used.

## 2.1 Simple Function (No Inputs)

```{r}
# Define a simple function
hello <- function() {
  print("hello!")
}

# Call the function
hello()
```

**Explanation**:\
- `hello()` prints `"hello!"` every time it’s called.\
- It has no arguments, so it always behaves the same.

------------------------------------------------------------------------

## 2.2 Function with One Input

```{r}
helloyou <- function(name) {
  print(paste("hello", name, "!"))
}

# Provide an argument
helloyou("Piotr")
```

**Explanation**:\
- `helloyou` takes one parameter `name`.\
- It uses `paste()` to combine strings and prints the result.\
- If you call `helloyou("Piotr")`, it prints `"hello Piotr !"`.

------------------------------------------------------------------------

## 2.3 Adding Two Numbers

```{r}
add_num <- function(num1, num2) {
  print(num1 + num2)
}

add_num(5, 10)
```

**Explanation**:\
- `add_num` accepts two numbers and prints their sum.\
- `add_num(5, 10)` prints `15`.

------------------------------------------------------------------------

## 2.4 Functions with Default Values

```{r}
hello_someone <- function(name = "Janusz") {
  print(paste("Hello", name))
}

# Uses default
hello_someone()

# Overwrite default
hello_someone("Piotr")
```

**Explanation**:\
- If no argument is supplied, `name` defaults to `"Janusz"`.\
- You can override the default by calling, for instance, `hello_someone("Piotr")`.

------------------------------------------------------------------------

## 2.5 Returning Values

Instead of printing the output directly, functions often **return** a value for further use.

```{r}
formal <- function(name = "Piotr", title = "Sir") {
  return(paste(title, name))
}

# Calling the function
formal()                 # "Sir Piotr"
formal("Isaac Newton")   # "Sir Isaac Newton"
```

**Explanation**:\
- `return()` explicitly specifies what the function should output.\
- If you omit `return(...)`, R returns the value of the last expression by default.

### Assigning Return Values

```{r}
var <- formal("Marie Curie", "Ms.")
var
```

-   The returned string `"Ms. Marie Curie"` is saved into `var`.

------------------------------------------------------------------------

# 3. Variable Scope

**Scope** refers to where a variable is **accessible** in your code.

-   **Local Variables**: Defined within a function; accessible only inside that function.\
-   **Global Variables**: Defined outside any function; accessible from anywhere in the R session.

## 3.1 Local vs. Global Scope

```{r}
pow_two <- function(input) {
  res <- input ^ 2
  return(result)
}

pow_two(4)    # returns 16

# Attempt to access local variables
res  # Error: object 'res' not found
input   # Error: object 'input' not found
```

**Explanation**:\
- `res` and `input` exist only inside `pow_two()`. They do not exist in the global environment.

------------------------------------------------------------------------

## 3.2 Accessing Global Variables

```{r}
v <- "I'm global v"
stuff <- "I'm global stuff"

fun <- function(stuff) {
  print(v)    # uses global v
  stuff <- "Reassign stuff inside func"
  print(stuff)
}

fun(stuff)
stuff  # Still "I'm global stuff" globally
```

**Explanation**:\
- `v` and `stuff` are global.\
- The local parameter `stuff` in `fun()` does not change the global `stuff`.

### Caution: Modifying Global Variables

You *can* change global vars from inside a function with `<<-`, but it’s generally discouraged:

```{r}
var <- 5

modify_var <- function() {
  var <<- var * 2
}
modify_var()
var  # now 10
```

Use this feature sparingly to avoid confusion.

------------------------------------------------------------------------

# 4. Additional Function Concepts

## 4.1 Vectorized Functions

R’s strength lies in **vectorized** operations. Rather than looping manually, you can perform operations on entire vectors at once.

```{r}
numbers <- 1:5
squared <- numbers ^ 2
squared
```

-   This squares each element, no explicit loop needed, typically faster and cleaner.

------------------------------------------------------------------------

## 4.2 Anonymous Functions

Sometimes you need a **function without a name**, for short, one-off tasks (especially in `apply`, `lapply`, `sapply`, etc.):

```{r}
numbers <- 1:5
squared <- sapply(numbers, function(x) x^2)
squared
```

**Explanation**:\
- The anonymous function `function(x) x^2` is defined inline and used by `sapply`.

------------------------------------------------------------------------

## 4.3 Higher-Order Functions

A **higher-order** function takes a function as an argument or returns a function. This can produce more abstract, reusable code.

```{r}
apply_func <- function(func, num) {
  return(func(num))
}

double_num <- function(x) {
  return(x * 2)
}

result <- apply_func(double_num, 5)  # 10
result
```

------------------------------------------------------------------------

# 5. Exercises

Below are practice exercises to strengthen your understanding of functions in R. You may include or remove **Proposed Solutions** as you see fit.

------------------------------------------------------------------------

### Exercise 1: Product of Two Integers

**Question**\
Create a function that returns the product of two integers.

```{r}
# Your code here:

adding <- function(x,y) {
  return (x+y)
}

print(adding(2,4))
```

------------------------------------------------------------------------

### Exercise 2: Check Integer in Vector

**Question**\
A function that takes an integer and a vector of integers, returning `TRUE` if the integer is present, otherwise `FALSE`.

```{r}
# Your code here:

is_present <- function(num, vec) {
  return(num %in% vec)
}

is_present(3, c(1, 2, 3, 4, 5))  
is_present(10, c(1, 2, 3, 4, 5)) 

```

------------------------------------------------------------------------

### Exercise 3: Count Occurrences in Vector

**Question**\
A function that takes an integer and a vector of integers, returning how many times the integer appears.

```{r}
# Your code here:

how_many <- function(num, vec) {
  return(sum(vec == num)) 
}
  
how_many(3, c(1, 2, 3, 3, 4, 5, 3))
how_many(10, c(1, 2, 3, 4, 5)) 
```

------------------------------------------------------------------------

### Exercise 4: Minimum Bars of Aluminum

**Question**\
We have small bars (1kg) and big bars (5kg). Given an integer of requested kilograms, return the **minimum** number of bars needed. For example: - 6 kg → 2 bars (5 + 1) - 17 kg → 5 bars (5 + 5 + 5 + 1 + 1)

```{r}
# Your code here:

bars <- function(kg) {
  small <- kg %% 5
  big <- (kg - small) / 5
  
  return (c(big,small))
}

bars(19)
bars(16)
```

------------------------------------------------------------------------

### Exercise 5: Check if an Integer is Prime

**Question**\
Create a function that returns `TRUE` if an integer is prime, else `FALSE`.

```{r}
# Your code here:
is_prime <- function(x) {
  if (x < 2) return(FALSE)
  return(!(as.logical(x %% 2)))
}

is_prime(5)
is_prime(1)
is_prime(6)
```

------------------------------------------------------------------------

### Exercise 6: Factorial of a Number

**Question**\
Create a function that calculates the factorial of a number using a `for` loop.

```{r}
# Your code here:

factorial <- function(x) {
  r <- 1
  for (i in 1:x) {
    r <- r * i 
  }
  return (r)
}

factorial(4)
```

------------------------------------------------------------------------

### Exercise 7: Greeting Function with Multiple Inputs

**Question**\
A function that takes `first_name` and `last_name`, returning a personalized greeting.

```{r}
# Your code here:

greetings <- function(first_name, last_name) {
  return (paste("Hello", first_name, last_name))
}

greetings("A", "Z")
```

------------------------------------------------------------------------

### Exercise 8: Area of a Circle

**Question**\
A function that calculates the area of a circle given its radius `r`. The formula is $\pi r^2$.

```{r}
# Your code here:

area <- function(r) {
  return (pi *r^2)
}

area(5)
```

------------------------------------------------------------------------

### Exercise 9: Celsius to Fahrenheit

**Question**\
A function that converts a temperature from Celsius to Fahrenheit using the formula $F = C \times \frac{9}{5} + 32$.

```{r}
# Your code here:

convert <- function(temp_c) { 
  return(temp_c * (9/5) + 32)
}

convert(30)
convert(20)
convert(10)
convert(0)
```

------------------------------------------------------------------------

### Exercise 10: Length of a String

**Question**\
A function that takes a string and returns its length using `nchar()`.

```{r}
# Your code here:

len <- function(s) {
  return(nchar(s))
}

len('ABC')
```

------------------------------------------------------------------------

# 6. Summary

In this notebook, we covered:

1.  **Defining Functions** with various arguments (including default values)\
2.  **Returning Values** vs. printing them\
3.  **Variable Scope** (local vs. global)\
4.  Additional concepts (vectorized operations, anonymous functions, higher-order functions)

**Key Takeaways**:

-   Functions in R are declared with `function(...) { ... }`.\
-   Use `return(...)` to explicitly give back a value.\
-   Respect **scope**: local variables are internal, global variables are external.\
-   Vectorized and higher-order functions can simplify code.

**Next Steps**:

1.  Practice writing more functions to solve different tasks.\
2.  Explore **recursive functions**, **closures**, and **function factories** for more advanced usage.\
3.  Learn about **package development** to organize related functions into distributable libraries.

------------------------------------------------------------------------

# References

-   [R Documentation: `function`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/function)\
-   [R for Data Science (Functions chapter)](https://r4ds.had.co.nz/)\
-   [Advanced R by Hadley Wickham (Functions section)](https://adv-r.hadley.nz/functions.html)\
-   [Understanding Function Scope in R](https://www.statmethods.net/advstats/scope.html)

------------------------------------------------------------------------

# Additional Resources

-   **RStudio Cheatsheets**:\
    [RStudio Cheatsheets](https://rstudio.com/resources/cheatsheets/)\
-   **Swirl - Learn R in R**:\
    [Interactive swirl tutorials](https://swirlstats.com/students.html)
